---
title: "Lab 3. Reserve Planning"
author: "Joe DeCesaro"
date: "2/7/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
## R packages {#r-packages}
```{r}
if (!require("librarian")){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  assertthat, BiocManager, dplyr, gridExtra, here, mapview, 
  prioritizr, prioritizrdata, 
  raster, remotes, rgeos, rgdal, scales, sf, sp, stringr,
  units)
if (!require("lpsymphony")){
  BiocManager::install("lpsymphony")
  library(lpsymphony)
}
```

## Data {#data-setup}
The data for this workshop are available online. To download the data, please  [click on this link](https://github.com/prioritizr/massey-workshop/raw/main/data.zip) and save the `data.zip` file to your computer. Next, unzip the `data.zip` file. You should now have a new folder on your computer called `"data"` which contains the data files (e.g. `pu.shp`).

```{r}
dir_data <- here("data/")
pu_shp   <- file.path(dir_data, "pu.shp")
pu_url   <- "https://github.com/prioritizr/massey-workshop/raw/main/data.zip"
pu_zip   <- file.path(dir_data, basename(pu_url))
vegetation_tif <- file.path(dir_data, "vegetation.tif")

dir.create(dir_data, showWarnings = F, recursive = T)
if (!file.exists(pu_shp)){
  download.file(pu_url, pu_zip)
  unzip(pu_zip, exdir = dir_data)
  dir_unzip   <- file.path(dir_data, "data")
  files_unzip <- list.files(dir_unzip, full.names = T)
  file.rename(
    files_unzip, 
    files_unzip %>% str_replace("prioritizr/data", "prioritizr"))
  unlink(c(pu_zip, dir_unzip), recursive = T)
}
```

# Data
## Data import
Now that we have the downloaded dataset, we will need to import it into our R session. Specifically, this data was obtained from the “Introduction to Marxan” course and was originally a subset of a larger spatial prioritization project performed under contract to Australia’s Department of Environment and Water Resources. It contains vector-based planning unit data (pu.shp) and the raster-based data describing the spatial distributions of 32 vegetation classes (vegetation.tif) in southern Tasmania, Australia. Please note this dataset is only provided for teaching purposes and should not be used for any real-world conservation planning. We can import the data into our R session using the following code.

```{r}
# import planning unit data
pu_data <- as(read_sf(pu_shp), "Spatial")

# format columns in planning unit data
pu_data$locked_in <- as.logical(pu_data$locked_in)
pu_data$locked_out <- as.logical(pu_data$locked_out)

# import vegetation data
veg_data <- stack(vegetation_tif)
```

## Planning unit data

The planning unit data contains spatial data describing the geometry for each planning unit and attribute data with information about each planning unit (e.g. cost values). Let's investigate the `pu_data` object. The attribute data contains `r ncol(pu_data)` columns with contain the following information:

* `id`: unique identifiers for each planning unit
* `cost`: acquisition cost values for each planning unit (millions of Australian dollars).
* `status`: status information for each planning unit (only relevant with Marxan)
* `locked_in`: logical values (i.e. `TRUE`/`FALSE`) indicating if planning units are covered by protected areas or not.
* `locked_out`: logical values (i.e. `TRUE`/`FALSE`) indicating if planning units cannot be managed as a protected area because they contain are too degraded.

```{r}
# print a short summary of the data
print(pu_data)

# plot the planning unit data
plot(pu_data)
```

```{r, eval = FALSE}
# plot an interactive map of the planning unit data
mapview(pu_data)
```

```{r, out.width = "60%"}
# print the structure of object
str(pu_data, max.level = 2)

# print the class of the object
class(pu_data)

# print the slots of the object
slotNames(pu_data)

# print the coordinate reference system
print(pu_data@proj4string)

# print number of planning units (geometries) in the data
nrow(pu_data)

# print the first six rows in the data
head(pu_data@data)

# print the first six values in the cost column of the attribute data
head(pu_data$cost)

# print the highest cost value
max(pu_data$cost)

# print the smallest cost value
min(pu_data$cost)

# print average cost value
mean(pu_data$cost)

# plot a map of the planning unit cost data
spplot(pu_data, "cost")
```

```{r, eval = FALSE}
# plot an interactive map of the planning unit cost data
mapview(pu_data, zcol = "cost")
```

Now, you can try and answer some questions about the planning unit data.

```{block2, type="rmdquestion"}
1. How many planning units are in the planning unit data?
2. What is the highest cost value?
3. Is there a spatial pattern in the planning unit cost values (hint: use `plot` to make a map)?
```

1. `r nrow(pu_data)`
2.`r max(pu_data$cost)` 
3. There is a spatial patter in the planning unit cost values with the least expensive land generally in the northeast, the most expensive land generally in the north, and the middle cost land found in the middle and western parts.

## Vegetation data

The vegetation data describe the spatial distribution of `r n_features` vegetation classes in the study area. This data is in a raster format and so the data are organized using a grid comprising square grid cells that are each the same size. In our case, the raster data contains multiple layers (also called "bands") and each layer has corresponds to a spatial grid with exactly the same area and has exactly the same dimensionality (i.e. number of rows, columns, and cells). In this dataset, there are `r n_features` different regular spatial grids layered on top of each other -- with each layer corresponding to a different vegetation class -- and each of these layers contains a grid with `r raster::nrow(veg_data)` rows, `r raster::ncol(veg_data)` columns, and `r nrow(veg_data) * ncol(veg_data)` cells. Within each layer, each cell corresponds to a `r raster::xres(veg_data)/1000` by `r raster::yres(veg_data)/1000` km square. The values associated with each grid cell indicate the (one) presence or (zero) absence of a given vegetation class in the cell.

Let's explore the vegetation data.

```{r "explore feature data"}
# print a short summary of the data
print(veg_data)

# plot a map of the 20th vegetation class
plot(veg_data[[20]])
```

```{r, eval = FALSE}
# plot an interactive map of the 20th vegetation class
mapview(veg_data[[20]])
```

```{r "preview feature data"}
# print number of rows in the data
nrow(veg_data)

# print number of columns  in the data
ncol(veg_data)

# print number of cells in the data
ncell(veg_data)

# print number of layers in the data
nlayers(veg_data)

# print  resolution on the x-axis
xres(veg_data)

# print resolution on the y-axis
yres(veg_data)

# print spatial extent of the grid, i.e. coordinates for corners
extent(veg_data)

# print the coordinate reference system
print(veg_data@crs)

# print a summary of the first layer in the stack
print(veg_data[[1]])

# print the value in the 800th cell in the first layer of the stack
print(veg_data[[1]][800])

# print the value of the cell located in the 30th row and the 60th column of
# the first layer
print(veg_data[[1]][30, 60])

# calculate the sum of all the cell values in the first layer
cellStats(veg_data[[1]], "sum")

# calculate the maximum value of all the cell values in the first layer
cellStats(veg_data[[1]], "max")

# calculate the minimum value of all the cell values in the first layer
cellStats(veg_data[[1]], "min")

# calculate the mean value of all the cell values in the first layer
cellStats(veg_data[[1]], "mean")
```

Now, you can try and answer some questions about the vegetation data.

```{block2, type="rmdquestion"}
1. What part of the study area is the 13th vegetation class found in (hint: make a map)? For instance, is it in the south-eastern part of the study area?
2. What proportion of cells contain the 12th vegetation class?
3. Which vegetation class is the most abundant (i.e. present in the greatest number of cells)?
```

1. Predominately the northeastern and western parts
2. 
3. 
```{r}
# mapview(veg_data[[13]])
```

# Gap analysis

## Introduction

Before we begin to prioritize areas for protected area establishment, we should first understand how well existing protected areas are conserving our biodiversity features (i.e. native vegetation classes in Tasmania, Australia). This step is critical: we cannot develop plans to improve conservation of biodiversity if we don't understand how well existing policies are currently conserving biodiversity! To achieve this, we can perform a "gap analysis". A gap analysis involves calculating how well each of our biodiversity features (i.e. vegetation classes in this exercise) are represented (covered) by protected areas. Next, we compare current representation by protected areas of each feature (e.g. 5% of their spatial distribution covered by protected areas) to a target threshold (e.g. 20% of their spatial distribution covered by protected areas). This target threshold denotes the minimum amount (e.g. minimum proportion of spatial distribution) that we need of each feature to be represented in the protected area system. Ideally, targets should be based on an estimate of how much area or habitat is needed for ecosystem function or species persistence. In practice, targets are generally set using simple rules of thumb (e.g. 10% or 20%), policy (17%; https://www.cbd.int/sp/targets/rationale/target-11) or standard practices (e.g. setting targets for species based on geographic range size) [@r1; @r2].

## Feature abundance

Now we will perform some preliminary calculations to explore the data. First, we will calculate how much of each vegetation feature occurs inside each planning unit (i.e. the abundance of the features). To achieve this, we will use the `problem` function to create an empty conservation planning problem that only contains the planning unit and biodiversity data. We will then use the `feature_abundances` function to calculate the total amount of each feature in each planning unit.

```{r}
# create prioritizr problem with only the data
p0 <- problem(pu_data, veg_data, cost_column = "cost")

# print empty problem,
# we can see that only the cost and feature data are defined
print(p0)

# calculate amount of each feature in each planning unit
abundance_data <- feature_abundances(p0)

# print abundance data
print(abundance_data)
```


```{r}
# note that only the first ten rows are printed,
# this is because the abundance_data object is a tibble (i.e. tbl_df) object
# and not a standard data.frame object
print(class(abundance_data))

# we can print all of the rows in abundance_data like this
print(abundance_data, n = Inf)
```

The `abundance_data` object contains three columns. The `feature` column contains the name of each feature (derived from `names(veg_data)`), the `absolute_abundance` column contains the total amount of each feature in all the planning units, and the `relative_abundance` column contains the total amount of each feature in the planning units expressed as a proportion of the total amount in the underlying raster data. Since all the raster cells containing vegetation overlap with the planning units, all of the values in the `relative_abundance` column are equal to one (meaning 100%). Now let's add a new column with the feature abundances expressed in area units (i.e. km^2^).

```{r}
# add new column with feature abundances in km^2
abundance_data$absolute_abundance_km2 <-
  (abundance_data$absolute_abundance * prod(res(veg_data))) %>%
  set_units(m^2) %>%
  set_units(km^2)

# print abundance data
print(abundance_data)
```

Now let's explore the abundance data.

```{r}
# calculate the average abundance of the features
mean(abundance_data$absolute_abundance_km2)

# plot histogram of the features' abundances
hist(abundance_data$absolute_abundance_km2, main = "Feature abundances")

# find the abundance of the feature with the largest abundance
max(abundance_data$absolute_abundance_km2)

# find the name of the feature with the largest abundance
abundance_data$feature[which.max(abundance_data$absolute_abundance_km2)]
```

Now, try to answer the following questions.

```{block2, type="rmdquestion"}
1. What is the median abundance of the features (hint: `median`)?
2. What is the name of the feature with smallest abundance?
3. How many features have a total abundance greater than 100 km^2 (hint: use `sum(abundance_data$absolute_abundance_km2 > set_units(threshold, km^2)` with the correct `threshold` value)?

```

1. `r median(abundance_data$absolute_abundance_km2)`
2. `r abundance_data$feature[which.min(abundance_data$absolute_abundance_km2)]`
3. 6

## Feature representation

After calculating the total amount of each feature in the planning units (i.e. the features' abundances), we will now calculate the amount of each feature in the planning units that are covered by protected areas (i.e. feature representation by protected areas). We can complete this task using the `eval_feature_representation_summary()` function. This function requires (i) a conservation problem object with the planning unit and biodiversity data and also (ii) an object representing a solution to the problem (i.e an object in the same format as the planning unit data with values indicating if the planning units are selected or not).

```{r}
# create column in planning unit data with binary values (zeros and ones)
# indicating if a planning unit is covered by protected areas or not
pu_data$pa_status <- as.numeric(pu_data$locked_in)

# calculate feature representation by protected areas
repr_data <- eval_feature_representation_summary(p0, pu_data[, "pa_status"])

# print feature representation data
print(repr_data)
```

Similar to the abundance data before, the `repr_data` object contains three columns. The `feature` column contains the name of each feature, the `absolute_held` column shows the total amount of each feature held in the solution (i.e. the planning units covered by protected areas), and the `relative_held` column shows the proportion of each feature held in the solution (i.e. the proportion of each feature's spatial distribution held in protected areas). Since the `absolute_held` values correspond to the number of grid cells in the `veg_data` object with overlap with protected areas, let's convert them to area units (i.e. km^2^) so we can report them.

```{r}
# add new column with the areas represented in km^2
repr_data$absolute_held_km2 <-
  (repr_data$absolute_held * prod(res(veg_data))) %>%
  set_units(m^2) %>%
  set_units(km^2)

# print representation data
print(repr_data)
```
Now let's investigate how well the species are represented.

```{block2, type="rmdquestion"}
1. What is the average proportion of the features held in protected areas (hint: use `mean(table$relative_held)` with the correct `table` name)?
2. If we set a target of 10% coverage by protected areas, how many features fail to meet this target (hint: use `sum(table$relative_held >= target_value)`  with the correct `table` name)?
3. If we set a target of 20% coverage by protected areas, how many features fail to meet this target?
4. Is there a relationship between the total abundance of a feature and how well it is represented by protected areas (hint: `plot(abundance_data$absolute_abundance ~ repr_data$relative_held)`)?

```





